---
title: "Class 09: Halloween Candy Mini Project"
author: "Helen Le (PID: A16300695)"
format: pdf
---

Today is Halloween! We will apply lots of the analysis methods and R graphics approaches to find out all about typical Halloween candy.

```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy <- read.csv(candy_file, row.names=1)
head(candy)
```

**Q1. How many different candy types are in this dataset?**

```{r}
dim(candy)
```
There are 85 different candy types in this dataset.

**Q2. How many fruity candy types are in the dataset?**

```{r}
sum(candy["fruity"])
# or, sum(candy$fruity)
```
There are 38 fruity candy types.

I can convert the 1 and 0 values to be TRUE or FALSE and use that to extract the tpe of candy I want. For example:
```{r}
candy[as.logical(candy$chocolate),]
```

# 2. What's your favorite candy

**Q3. What is your favorite candy in the dataset and what is it’s winpercent value?**

```{r}
candy["Reese's Peanut Butter cup", ]$winpercent
```
My favorite candy has a win percent of 84.18%.

**Q4. What is the winpercent value for “Kit Kat”?**

```{r}
candy["Kit Kat",]$winpercent
```
Kit Kat has a winpercent value of 76.77%. 

**Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?**

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```
Tootsie Roll Snack Bars have a winpercent value of 49.65%.

The "skimr" package has a function that helps to give a quick overview of a given dataset.
```{r}
library("skimr")
skim(candy)
```

**Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?**
The winpercent variable looks to be on a different scale to the majority of the other columns.


**Q7. What do you think a zero and one represent for the candy$chocolate column?**
A zero and one likely represents logicals for the candy types, determining whether it is TRUE or FALSE that the candy type is considered chocolate or not.

**Q8. Plot a histogram of winpercent values**
```{r}
library(ggplot2)
```

```{r}
hist(candy$winpercent)
```


```{r}
ggplot(candy) +
  aes(winpercent) +
  geom_histogram(binwidth=10)
```


**Q9. Is the distribution of winpercent values symmetrical?**
No, they're not symmetrical.

**Q10. Is the center of the distribution above or below 50%?**
The center of distribution is below 50%.

**Q11. On average is chocolate candy higher or lower ranked than fruit candy?**
```{r}
choco.win <- candy$winpercent[as.logical(candy$chocolate)]
fruit.win <- candy$winpercent[as.logical(candy$fruity)]

mean(choco.win)
mean(fruit.win)
```
Another way to do this is:
```{r}
choc.inds <- as.logical(candy$chocolate)
choc.win <- candy[choc.inds, "winpercent"]
choc.win
```

On average, chocolate candy is ranked higher than fruit candy.

**Q12. Is this difference statistically significant?**
```{r}
t.test(choco.win, fruit.win)
```
This difference is statistically significant as the two-sample t-test yielded a low p-value of 2.871e-08.

# 3. Overall Candy Rankings

**Q13. What are the five least liked candy types in this set?**
```{r}
head(candy[order(candy$winpercent),], n=5)
```

```{r}
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```

The five least liked candy types in this set are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.

**Q14. What are the top 5 all time favorite candy types out of this set?**
```{r}
head(candy[order(candy$winpercent, decreasing=TRUE),], n=5)
```
The top 5 all time favorite candy types out of this set are Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, and Snickers.

**Q15. Make a first barplot of candy ranking based on winpercent values.**
```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

**Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?**
```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

Let's add a color vector to our plot.
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

**Q17. What is the worst ranked chocolate candy?**
The worst ranked chocolate candy is Sixlets.

**Q18. What is the best ranked fruity candy?**
The best ranked fruity candy is Starburst.

# 4. Taking a look at pricepercent

```{r}
library(ggrepel)

# Plot of price vs. winpercent
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 8)
```
`geom_text_repel()` moves around the labels so that the text becomes more legible on the plot.


**Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?**
Reese's Miniatures is the candy type that is ranked the highest in terms of winpercent for the least money.

**Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?**
```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

The top 5 most expensive candy types in the data set are Nik L Nip, Ring pop, Nestle Smarties, Hershey's Krackel, and Hershey's Milk Chocolare. Of these, Nik L Nip is the least popular.

# 5. Exploring Correlation Structure

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

**Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?**
Chocolate and fruity candy types are anti-correlated. (They are negatively associated with each other.) This makes sense since you don't often see chocolatey fruity candies.

**Q23. Similarly, what two variables are most positively correlated?**
Chocolate and bar candy types are the most positively correlated. This makes sense since chocolate candies come in a bar form.

# 6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

Let's try to plot the same thing using ggplot2.
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

Let's add labels to this plot:
```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
The max.overlaps argument can be changed to allow more overlapping labels.

Or, plotly can be used to create a more interactive plot that will show the data labels when it is moused over.
*Note: This only works in HTML format*
```{r}
library(plotly)
```
The following interactive plot is commented out since it doesn't work in pdf format.
```{r}
# ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

**Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?**
Fruity,hard, and pluribus candy types are picked up strongly by PC1 in the positive direction. This makes sense as they are commonly associated with each other. They are also anti-correlated with the other variables, like chocolate and bar.