---
title: "Class 12: RNASeq Galaxy"
format: pdf
---

# Section 1. Proportion of G/G in a population

Downloaded a CSV file from Ensemble <
- saved as mxl.csv in our directory

Here, we read this CSV file:
```{r}
mxl <- read.csv("mxl.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

Now, let's look at a different population. I picked the GBR.
```{r}
gbr <- read.csv("gbr.csv")
```

Find proportion of G|G:
```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population.

Let's now dig into this further.

# Section 4: Population Scale Analysis [HOMEWORK]

One sample is obviously not enough to know what is happening in a population. You are 
interested in assessing genetic differences on a population scale. So, you processed about 
~230 samples and did the normalization on a genome level. Now, you want to find whether 
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 
expression.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```
**Q13. The sample sizes for the A/A, A/G, and G/G genotypes are 108, 233, 121, respectively.**


```{r}
library(ggplot2)
```

Let's make a boxplot
```{r}
ggplot(expr) + aes(x=geno, y=exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```

**Q14. There is relatively more expression of the A/A genotype than the G/G genotype, as displayed in the plot. Yes, the SNP does effect the expression of ORMDL3.**